Searching for long-period transiting planets in the \kepler\ light curves using supervised classification

### 2015-04-07 document/ms.tex

\newcommand{\figref}[1]{\ref{fig:#1}}
\newcommand{\Fig}[1]{\figurename~\figref{#1}}
\newcommand{\fig}[1]{\Fig{#1}}
\newcommand{\figlabel}[1]{\label{fig:#1}}
\newcommand{\Tab}[1]{Table~\ref{tab:#1}}
\newcommand{\tab}[1]{\Tab{#1}}
\newcommand{\tablabel}[1]{\label{tab:#1}}
\newcommand{\Eq}[1]{Equation~(\ref{eq:#1})}
\newcommand{\eq}[1]{\Eq{#1}}
\newcommand{\eqalt}[1]{Equation~\ref{eq:#1}}
\newcommand{\eqlabel}[1]{\label{eq:#1}}
\newcommand{\sectionname}{Section}
\newcommand{\Sect}[1]{\sectionname~\ref{sect:#1}}
\newcommand{\sect}[1]{\Sect{#1}}
\newcommand{\sectalt}[1]{\ref{sect:#1}}
\newcommand{\App}[1]{Appendix~\ref{sect:#1}}
\newcommand{\app}[1]{\App{#1}}
\newcommand{\sectlabel}[1]{\label{sect:#1}}

@@

Before the launch of the \kepler\ Mission, \citet{Yee:2008} predicted the
yield of single transit events based on the planet occurrence rates estimated
based on small catalog of radial velocity discoveries \citep{Butler:2006}.
Using these early results and the pre-launch specifications of the \kepler\
Mission, \citet{Yee:2008} predicted that \kepler\ would discover $\sim 6$
single transit events in the full dataset.
Now that the Mission is complete and we have better estimates of the
occurrence rate and distribution of planets \citep[for example][]{Dong:2013,
Petigura:2013, Foreman-Mackey:2014, Dressing:2015}, we can update the
predicted yield for single transit events in the \kepler\ light curves.
To make this estimate, we extrapolate the distribution of large planets on
relatively short orbits out to longer periods, taking detection efficiency
and the survey and targeting properties into account.

@@

For the purposes of this discussion, we assume an approximate simple detection
efficiency model with three contributions: the geometric transit probability,
the temporal transit probability, and signal-to-noise ratio threshold of the
search technique.
Assuming circular orbits, the geometric transit probability is given by
\citep{Winn:2010}
\begin{eqnarray}
Q_k^\mathrm{(geom)}(\rp,\,\period) &=& \frac{R_k}{a} \\
&=& \left( \frac{4\,\pi^2}{G\,M_k} \right)^{1/3} \, R_k \, \period^{-2/3}
\end{eqnarray}
where $R_k$ and $M_k$ are the radius and mass of the star $k$ respectively.
The eccentricity distribution of these long-period planets will affect this
transit probability \citep{Kipping:2014} but this simple prescription should
be sufficient for a rough estimate.
For long-period orbits, the temporal transit probability will be given by
\begin{eqnarray}
Q_k^\mathrm{(time)}(\rp,\,\period) &=& \frac{T_k}{\period}
\end{eqnarray}
where $T_k$ is the total time that \kepler\ spent observing the star $k$.
Finally, the detection threshold depends on the detailed sensitivity of the
search procedure but we will approximate it as a simple step function in
signal-to-noise ratio of the transit.
This contribution will be given approximately by
\begin{eqnarray}
Q_k^\mathrm{(detect)}(\rp,\,\period) &=& \left\{\begin{array}{ll}
1 & \mathrm{if}\,\left(\rp/R_k\right)^2 > f\,\sigma_k \\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
where $\sigma_k$ is an estimate of the noise in light curve of star $k$ and
$f$ is the detection threshold for the method.

Combining the detection efficiency components, the integral from \eq{nk}
becomes
\begin{eqnarray}
N_k &=& \left( \frac{4\,\pi^2}{G\,M_k} \right)^{1/3} \, R_k \, T_k
    \int_{\rp_\mathrm{min}} ^{\rp_\mathrm{max}} \frac{\dd\rp}{\rp}
    \int_{\period_\mathrm{min}} ^{\period_\mathrm{max}}
        \period^{-8/3}\,\rate(\rp,\,\period) \dd\period
\end{eqnarray}
where all the integration limits are set by the target parameter space.
Because of the detection probability threshold, $\rp_\mathrm{min}$ can be no
smaller than $R_k\,\sqrt{f\,\sigma_k}$.
\citet{Dong:2013} used the catalog of short period transiting planets found
by \kepler\ to constrain a model for the occurrence rate of large planets of
the form
\begin{eqnarray}
\rate(\rp,\,\period) &=& C\,\left(\frac{\period}{10\,\mathrm{d}}\right)^\beta
\end{eqnarray}
in a set of radius bins.

Assuming the stellar parameters provided by the NASA Exoplanet
Archive\footnote{\dfmtodo{URL for kepler stellar table; downloaded on SOME
DATE.}} \dfmtodo{cite Huber} and approximating the stellar noise using the
15-hour CDPP \dfmtodo{cite CDPP}, \dfmtodo{some figure} shows the extrapolated
number of transiting planets with $\period > 1500\,\mathrm{d}$ based on the
\citet{Dong:2013} power-law model.
For a detection threshold $f\sim10$, the expected number of single transit
events in the \kepler\ light curves is $\sim 120$.

@@

The \kepler\ Mission measured photometry for about 150,000 stars at half-hour
cadence for a baseline of over four years.
We aim to search these time series for single transits of long-period planets
and single eclipses of binary stars.
These data are made available on
MAST\footnote{\url{https://archive.stsci.edu/kepler/}} and, for each target,
we downloaded the full set of light curve files provided by Data Release 24
\citep{Thompson:2015}.
From these files, we extracted the PDC time series and split them into
``sections'' with no more than two contiguous missing or flagged data points.
The PDC light curves have been corrected for the instrumental effects caused
by the spacecraft using a data-driven model of the focal plane
\citep{Stumpe:2012, Smith:2012}.
Crucially, an attempt is also made by the PDC procedure to remove sharp
instrumental artifacts like ``sudden pixel sensitivity dropouts (SPSDs)''.

@@

\section{Demonstration}\sectlabel{demo}

@@

@@

>>>>

\newcommand{\figref}[1]{\ref{fig:#1}}
\newcommand{\Fig}[1]{\figurename~\figref{#1}}
\newcommand{\fig}[1]{\Fig{#1}}
\newcommand{\figlabel}[1]{\label{fig:#1}}
\newcommand{\Tab}[1]{Table~\ref{tab:#1}}
\newcommand{\tab}[1]{\Tab{#1}}
\newcommand{\tablabel}[1]{\label{tab:#1}}
\renewcommand{\eqref}[1]{\ref{eq:#1}}
\newcommand{\Eq}[1]{Equation~(\eqref{#1})}
\newcommand{\eq}[1]{\Eq{#1}}
\newcommand{\eqalt}[1]{Equation~\eqref{#1}}
\newcommand{\eqlabel}[1]{\label{eq:#1}}
\newcommand{\sectionname}{Section}
\newcommand{\Sect}[1]{\sectionname~\ref{sect:#1}}
\newcommand{\sect}[1]{\Sect{#1}}
\newcommand{\sectalt}[1]{\ref{sect:#1}}
\newcommand{\App}[1]{Appendix~\ref{sect:#1}}
\newcommand{\app}[1]{\App{#1}}
\newcommand{\sectlabel}[1]{\label{sect:#1}}

@@

Before the launch of the \kepler\ Mission, \citet{Yee:2008} predicted the
yield of single transit events based on the planet occurrence rates estimated
based on small catalog of radial velocity discoveries \citep{Butler:2006} and
a fit to their occurrence rate \citep{Cumming:2008}.
Using these early results and the pre-launch specifications of the \kepler\
Mission, \citet{Yee:2008} predicted that \kepler\ would discover $\sim 6$
single transit events in the full dataset.
Now that the Mission is complete and we have better estimates of the
occurrence rate and distribution of planets \citep[for example][]{Dong:2013,
Petigura:2013, Foreman-Mackey:2014, Dressing:2015}, we can update the
predicted yield for single transit events in the \kepler\ light curves.
To make this estimate, we extrapolate the distribution of large planets on
relatively short orbits out to longer periods, taking detection efficiency
and the survey and targeting properties into account.

@@

For the purposes of this discussion, we assume an approximate simple detection
efficiency model with three contributions: the geometric transit probability,
the temporal transit probability, and signal-to-noise ratio threshold of the
search technique.
Assuming circular orbits, the geometric transit probability is given by
\citep{Winn:2010}
\begin{eqnarray}\eqlabel{q-geom}
Q_k^\mathrm{(geom)}(\rp,\,\period) &=& \frac{R_k}{a} \\
&=& \left( \frac{4\,\pi^2}{G\,M_k} \right)^{1/3} \, R_k \, \period^{-2/3}
\end{eqnarray}
where $R_k$ and $M_k$ are the radius and mass of the star $k$ respectively.
The eccentricity distribution of these long-period planets will affect this
transit probability \citep{Kipping:2014} but this simple prescription should
be sufficient for a rough estimate.
For long-period orbits, the temporal transit probability will be given by
\begin{eqnarray}\eqlabel{q-time}
Q_k^\mathrm{(time)}(\rp,\,\period) &=& \frac{T_k}{\period}
\end{eqnarray}
where $T_k$ is the total time that \kepler\ spent observing the star $k$.
Finally, the detection threshold depends on the detailed sensitivity of the
search procedure but we will approximate it as a simple step function in
signal-to-noise ratio of the transit.
This contribution will be given approximately by
\begin{eqnarray}
Q_k^\mathrm{(detect)}(\rp,\,\period) &=& \left\{\begin{array}{ll}
1 & \mathrm{if}\,\left(\rp/R_k\right)^2 > f\,\sigma_k \\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
where $\sigma_k$ is an estimate of the noise in light curve of star $k$ and
$f$ is the detection threshold for the method.

Combining the detection efficiency components, the integral from \eq{nk}
becomes
\begin{eqnarray}\eqlabel{nk-2}
N_k &=& \left( \frac{4\,\pi^2}{G\,M_k} \right)^{1/3} \, R_k \, T_k
    \int_{\rp_\mathrm{min}} ^{\rp_\mathrm{max}} \frac{\dd\rp}{\rp}
    \int_{\period_\mathrm{min}} ^{\period_\mathrm{max}}
        \period^{-8/3}\,\rate(\rp,\,\period) \dd\period
\end{eqnarray}
where all the integration limits are set by the target parameter space.
Because of the detection probability threshold, $\rp_\mathrm{min}$ can be no
smaller than $R_k\,\sqrt{f\,\sigma_k}$.
\citet{Dong:2013} used the catalog of short period transiting planets found
by \kepler\ to constrain a model for the occurrence rate of large planets of
the form
\begin{eqnarray}
\rate(\rp,\,\period) &=& C\,\left(\frac{\period}{10\,\mathrm{d}}\right)^\beta
\end{eqnarray}
in a set of radius bins.
Using this model, the integral in \eq{nk-2} becomes
\begin{eqnarray}
N_k &=& \left( \frac{4\,\pi^2}{G\,M_k} \right)^{1/3} \, R_k \, T_k \,
    \sum_{j=1}^J \frac{C_j}{(10\,\mathrm{d})^{\beta_j}}\,
    \ln \left( \frac{\rp_{\mathrm{max},j}}{\rp_{\mathrm{min},j}} \right) \,
    \left[ \frac{{\period_\mathrm{min}}^{\beta_j - 5/3}}{5/3-\beta_j} \right]
\end{eqnarray}
where the sum is over the $J$ radial bins studied by \citet{Dong:2013}.
It's important to note that the model used by \citet{Dong:2013} is given in
base-10 logarithms so the units must be converted to natural logarithms as
appropriate.

Assuming the stellar parameters provided by the NASA Exoplanet
Archive\footnote{\dfmtodo{URL for kepler stellar table; downloaded on SOME
DATE.}} \dfmtodo{cite Huber} and approximating the stellar noise using the
15-hour CDPP \dfmtodo{cite CDPP}, \dfmtodo{some figure} shows the extrapolated
number of transiting planets with $\period > 1500\,\mathrm{d}$ based on the
\citet{Dong:2013} power-law model.
For a detection threshold $f\sim10$, the expected number of single transit
events in the \kepler\ light curves is $\sim 150$.

This results is much more optimistic than the pre-launch estimate from
\citet{Yee:2008} for a few reasons.
One effect is that, the \kepler\ Mission ran for more than four
years---longer than the fiducial Mission goal---and $\sim190,000$ stars were
targeted for nearly the full baseline instead of the original 100,000.
\dfmtodo{Explain other reasons.}

@@

The \kepler\ Mission measured photometry for about 190,000 stars at half-hour
cadence for a baseline of over four years.
We aim to search these time series for single transits of long-period planets
and single eclipses of binary stars.
These data are made available on
MAST\footnote{\url{https://archive.stsci.edu/kepler/}} and, for each target,
we downloaded the full set of light curve files provided by Data Release 24
\citep{Thompson:2015}.
From these files, we extracted the PDC time series and split them into
``sections'' with no more than two contiguous missing or flagged data points.
The PDC light curves have been corrected for the instrumental effects caused
by the spacecraft using a data-driven model of the focal plane
\citep{Stumpe:2012, Smith:2012}.
Crucially, an attempt is also made by the PDC procedure to remove sharp
instrumental artifacts like ``sudden pixel sensitivity dropouts (SPSDs)''.

@@

\section{Preliminary results}\sectlabel{demo}

\paragraph{A discovery}

@@

Even though only one transit is observed, it is still possible to estimate the
orbital period of these planet candidates using independent constraints on the
stellar properties.
This technique is similar in spirit to the ``photoeccentric effect''
\citep{Dawson:2012} and ``asterodensity profiling'' \citep{Kipping:2012}.
The fundamental principle is that, the transit duration is a measurement of
the instantaneous velocity of the planetary motion and this yields a
constraint on the orbital period when combining this with a measurement of the
stellar density and a prior on the eccentricity of the orbit.

In practice, we directly model the light curve as being generated by a planet
on an eccentric Keplerian orbit transiting a limb darkened star and
marginalize over all the orbital elements and physical parameters.
We take limb darkening and the finite exposure time into account
\citep{Mandel:2002, Kipping:2010, Kipping:2013} and use an empirical
Beta function prior on the eccentricity \citep{Kipping:2013a}.

@@

Based on our single transit discoveries, we now want to place a limit on
the occurrence rate of long-period planets.
Since we only have one discovery thus far, we'll make the simplifying
assumption that the rate is flat in log-period and log-radius in the range
$1000\,\mathrm{d} < \period < 5000\,\mathrm{d}$ and $5\,R_\odot <
20\,R_\odot$.
For this distribution, \citet{Foreman-Mackey:2014} derived the analytic
maximum likelihood result
\begin{eqnarray}
\rate &=& \frac{N_\mathrm{obs}}{\int Q(w)\dd w}
\end{eqnarray}
where $\rate$ is the occurrence rate per natural logarithm in period and
radius, $N_\mathrm{obs}$ is the number of detections, and the integral is over
the completeness measurements for every star that has been searched.
If we make the simplifying assumption that the detectability of a transit
depends on period only through Equations~(\eqref{q-geom})
and~(\eqref{q-time}), then this integral becomes
\begin{eqnarray}
\int Q(w)\dd w &=& \sum_{k=1}^K
    \left( \frac{4\,\pi^2}{G\,M_k} \right)^{1/3} \, R_k \, T_k \, Q_k \,
    \int \period^{-8/3}\,\rp^{-1} \dd\rp\dd\period \\
 &=& \sum_{k=1}^K
    \left( \frac{4\,\pi^2}{G\,M_k} \right)^{1/3} \, R_k \, T_k \, Q_k \,
    \ln \left(\frac{\rp_\mathrm{max}}{\rp_\mathrm{min}}\right)
    \frac{3}{5} \,
    \left[{\period_\mathrm{min}}^{-5/3} - {\period_\mathrm{max}}^{-5/3}\right]
\end{eqnarray}
where the sum is over all the targeted stars.
